---
title: "Simulation Exercise"
author: "Phil Renner"
date: "3/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Part 1: Simulating the Exponential Distribution

## Overview
In this project you will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. You will investigate the distribution of averages of 40 exponentials. Note that you will need to do a thousand simulations.

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials. You should

Show the sample mean and compare it to the theoretical mean of the distribution.
Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
Show that the distribution is approximately normal.
In point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.

## Simulations
I simulated the exponential distribution using rexp(40, 0.2), specifying 40 observations in each simulation with lambda = 0.2. I ran the simulation 1,000 times, creating vectors of 1,000 means and standard deviations.


```{r simulation}
set.seed(2021)
lambda <- 0.2
n_obs <- 40
mns = NULL
for (i in 1 : 1000) mns = c(mns, mean(rexp(n_obs, lambda)))

str(mns)
summary(mns)

stdvs = NULL
for (i in 1 : 1000) stdvs = c(stdvs, sqrt(var(rexp(n_obs, lambda))))

str(stdvs)
summary(stdvs)

```


## Analysis
## Question 1: Sample Mean vs Theoretical Mean

The theoretical mean for the exponential distribution is 1/lambda. In this case, lambda = 0.2 and the theoretical mean of each simulated distribution is 5.0. I plotted the mean of each simulation on a histogram. I also plotted the mean on the histogram. The mean of the means of the simulated distributions is `r mean(mns)`, so the simulated mean is very similar to the theoretical mean.

note add this to the plots to get normal
 geom_density(colour="blue", size=1) +
  scale_x_continuous(breaks=seq(mu-3,mu+3,1), limits=c(mu-3,mu+3))  +
  stat_function(fun = dnorm, args = list(mean = mu , sd = sd), colour = "red", size=1)


```{r means}

#plot histogram with mean as red line
g<- ggplot(data.frame(data = mns), aes(x=mns)) + geom_histogram(color="black", fill= "blue") +geom_vline(aes(xintercept = mean(mns)),col='red',size=2)+
  stat_function(fun = dnorm, args = list(mean = 5 , sd = 5), colour = "red", size=1)
g

```



## Analysis
## Question 2: Sample Variance vs Theoretical Variance
The theoretical standard deviation for the exponential distribution is 1/lambda/sqrt(n). We can calculate this theoretical standard deviation and compare it to the standard deviation of the simulated data. 

As you can see below, the observed and theoretical variance and standard deviation are nearly identical




```{r variance comparison}
#theoretical standard deviation
1/lambda/sqrt(n_obs)

#observed standard deviation
sd(mns)

#theoretical variance
(1/lambda/sqrt(n_obs))^2
  
#observed variance  
sd(mns)^2 


```



## Analysis
## Question 3: Distribution
The Central Limit Theorem holds that the means of a distribution are normally distributed. We will plot the means of the exponential functions. As you can see in the plot, the means are approximately normally distributed.

```{r normal}

#plot histogram with smoothed distribution and normal
g<- ggplot(data.frame(data = mns), aes(x=mns)) + geom_histogram(color="black", fill= "blue")+ 
 geom_density(aes(y=..count../5))  
 # stat_function(fun = dnorm, args = list(mean = 5 , sd = 5), colour = "red", size=1)
g

```



